---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# phobos

phobos is a process for identifying and ranking unknown mass features (MFs) in metabolomics data. The script annotates, ranks, and scores the unknown mass features, and performs data transformations common to the [Ingalls Lab](https://sites.google.com/view/anitra-ingalls). 

**Table of contents:** 

- [Confidence Level 1](https://github.com/IngallsLabUW/phobos#Confidence Level 1) 
- [Confidence Level 2](https://github.com/IngallsLabUW/phobos#Confidence Level 2) 
- [Confidence Level 3](https://github.com/IngallsLabUW/phobos#Confidence Level 3) 
- [Confidence Level 4](https://github.com/IngallsLabUW/phobos#Confidence Level 4) 


## Description

A common problem in metabolomics is how to handle unknown but persistently present mass features. The output of non-targeted runs produces an abundance of spectra that can be discretely separated and isolated, but often remain unidentified or identified with non-standard confidence. 

The phobos package prioritizes simplicity, efficiency, and flexibility in feature annotation. To encourage confidence consistency, phobos adheres to the proposed minimum reporting standards for chemical analysis as laid out in the Metabolomics Standards Initiative (MSI) [2007 paper](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3772505/pdf/nihms504189.pdf). In addition to utilizing the [Ingalls Lab Standards](https://github.com/IngallsLabUW/Ingalls_Standards) for foremost confidence annotations, phobos provides functionality for incorporating spectral comparison to third-party sources such as [MassBank of North America (MoNA)](https://mona.fiehnlab.ucdavis.edu/) and [KEGG](https://www.genome.jp/kegg/). The script uses equations from two papers: [Horai et al 2010](https://doi.org/10.1002/jms.1777) and [Tsugawa et al 2015](https://doi.org/10.1038/nmeth.3393).


In order to ensure that the MS2 (and MS1, rt, etc.) matching is accurate, we need to be comparing known data before we compare unknowns.
Horai et al. 2010 recommends taking a consensus of MS2 data for comparison, as reproducibility of MS2 data is notoriously difficult.

Therefore, we ran all of the Ingalls Standards 5 times (thanks Laura) and obtained five sets of MS2 data for each compound.
We then took a consensus of four of the runs to create "theoretical data"; i.e, something to compare the fifth run to for accuracy.
Because the fifth run is being compared to very similar runs, that removes variation that might interfere with comparisons.

Taking a consensus of the MS2 data:
- The data is scaled to an intensity of 100 and all intensities below 0.5 are removed.
- Fragments are grouped exclusively on mz, and assigned a group number for those that fall within a user-defined ppm range.
- Find clusters that are robust across files using hierarchical clustering in mz space, and cut the tree at a height that produces
  the most groups with 5 +/-1 fragments, or 10 +/- fragments to account for "superimposed lollipops" (ask Will).


```{r load data, include=FALSE}
library(tidyverse)
options(digits = 2) 
ingalls.standards <- read.csv("https://raw.githubusercontent.com/IngallsLabUW/Ingalls_Standards/master/Ingalls_Lab_Standards.csv",
                                                    stringsAsFactors = FALSE, header = TRUE) %>%
  select(compound_name = Compound_Name, HILIC_Mix, mz, rt = RT_minute, column = Column, z) %>%
  mutate(rt = rt * 60) %>%
  distinct()

example.experimental <- read_csv("example_data/Ingalls_Lab_Standards_MSMS.csv") %>%
  filter(str_detect(filename, "pos5|neg5")) %>%
  separate_rows(MS2, sep = "; ") %>%
  separate(MS2, into = c("mz", "int"), sep = ", ") %>%
  mutate(mz=as.numeric(mz)) %>%
  mutate(int=as.numeric(int)) %>%
  group_by(compound_name, filename, voltage) %>%
  mutate(int=int/max(int)*100) %>%
  arrange(desc(int)) %>%
  filter(int>1) %>%
  group_by(voltage, compound_name, filename) %>%
  summarise(MS2 = paste(mz, int, sep = ", ", collapse = "; ")) %>%
  mutate(z = ifelse(str_detect(filename, "pos"), 1, -1)) %>%
  mutate(HILIC_Mix = ifelse(str_detect(filename, "Mix1"), "Mix1", "Mix2")) %>%
  rowwise() %>%
  mutate(MS2 = paste(voltage, MS2, sep = "V ", collapse = ": ")) %>%
  left_join(ingalls.standards, by = c("compound_name", "z", "HILIC_Mix")) %>%
  select(compound_name, voltage, mz, rt, column, z, MS2) %>%
  drop_na()
example.experimental$MS2 <- sub("^([^;]+;[^;]+).*", "\\1", example.experimental$MS2)

example.theoretical <- read_csv("example_data/Ingalls_Lab_Standards_MSMS_consensed.csv") %>%
  mutate(z = ifelse(polarity == "pos", 1, -1)) %>%
  left_join(ingalls.standards, by = c("compound_name", "z")) %>%
  separate_rows(consensus_MS2, sep = ": ") %>%
  mutate(voltage = sub("\\V.*", "", consensus_MS2)) %>%
  select(compound_name, voltage, mz, rt, column, z, MS2 = consensus_MS2)
example.theoretical$MS2 <- sub("^([^;]+;[^;]+).*", "\\1", example.theoretical$MS2)
```

## Usage

phobos simplifies the identification process by transforming each individual MSI rank into a step for annotation. Users begin with a formatted data frame of experimental values, and compare those to theoretical values. With each successive function, a new column is added to the existing data frame with potential matches from each source.

Before any of the scripts can be run, the experimental data is transformed *by the user* into a standardized format. It is important to pay attention to both the column names and the class; the code will not run if they are not in the correct configuration!

Below is a list of the columns required to use phobos. The data frame must contain only these columns.

- "compound_name": Any identifier for a unique mass feature, character class.
- "voltage": Voltage at which the feature was run on the instrument, numeric class.
- "mz": The m/z value, numeric class.
- "rt": The retention time in seconds, numeric class.
- "column": The column the mass feature was run on, character class. There are two options for this variable: "HILIC" or "RP" (short for Reverse Phase).
- "z": The polarity, numeric class.
- "MS2": MS2 data for those compounds that have it, character class.
  - **Important**: The MS2 data must be in the filtered, scaled and concatenated format of "voltage V: mz, intensity;", as in the data frame below. For example purposes, the MS2 column below has been cut off at two instances of "mz, intensity;" but real data will have many more entries. Please see the "MS2voltageformat.R" script for assistance in getting your MS2 data to the correct form.

Below is an example of an experimental dataframe in the correct format.

```{r example experimental, echo=FALSE}
knitr::kable(head(example.experimental, 5))
```

---

### Confidence Level 1

In order to receive a Confidence Level 1 annotation, a mass feature must be compared to an authentic chemical standard (for this lab, the [Ingalls Standards sheet](https://github.com/IngallsLabUW/Ingalls_Standards/blob/master/Ingalls_Lab_Standards.csv)) using two independent and orthogonal checks. A mass feature must have a good match to a chemical standard on MS1, MS2, RT, column and z. 

The complete Ingalls Standards csv is located on the [Ingalls Standards Github](https://github.com/IngallsLabUW/Ingalls_Standards/blob/master/Ingalls_Lab_Standards.csv) and can be downloaded for the latest updated version. Like the experimental data, the theoretical data must be in the same specific format before comparison. 

**Tidy and setup** 

- Make sure all inputs (experimental and theoretical) are in a standardized format.

**Process**

- Apply comparison function to each row (each compound @ each voltage) of the experimetnal data frame. 
- Filter the potential matches on m/z by user-defined ppm error. 
- Match polarity and column. 
- Rowwise: 
  MS1 (m/z) and retention time (RT) Similarity Scores: exp(-0.5 * (((experimental value - theoretical value) / user-defined flexibility) ^ 2))
  MS2 cosine similarity. If both sets of MS2 values are present, calculate the similarity using the 2015 MSDial paper similarity equations.
  Total Similarity Score, according to which other similarity scores are present, also taken from 2015 MSDial paper.


Notes on CL1:
Prepping the data can be a big step. See the production of experimental.data for details.
Need to make an adjustment for "n consensus files", maybe not the current 4:6/9:11 being used for the testing.
  How many "intensity clusters" at mz points do we have? Now we're accounting
  for two intensity clusters, but it's possible we'd have three, four...
TODO: Randomize the 4:1 experimental:theoretical choices.
TODO: The flex arguments in the similarity score calculations are hard coded within the
  big ConfLevel1 function. Should we change that?
TODO: I have a progress bar wrapped around the MS2 Sim Score function but this could be better.
TODO: The med_sim_overall function is choosing some incorrect top choices. Going with the max
  sim score has fewer wrong choices, but still a decent amount.
TODO: Not so much for CL1, but in the later levels we might want to include the original theoretical data
  in the output columns. There's poor matching going on and it's difficult to see what is being matched.
TODO: Have "clean = TRUE" argument to drop the nested dataframe altogether.

---

### Confidence Level 2

Compounds that receive a confidence level of 2 are considered putatively annotated compounds; there are no available chemical reference standards, but they have good z, MS1 and MS2 spectral similarity with public/commercial spectral libraries. The current library used for confidence level 2 is [Massbank of North America](https://mona.fiehnlab.ucdavis.edu/).

The ready-to-compare MoNA spectra can be found on the shared Ingalls Lab Google Drive under Collaborative_Projects --> MARS_Project --> MoNA_RelationalSpreadsheets, but the public data can also be downloaded directly from the [public website](https://mona.fiehnlab.ucdavis.edu/downloads) for the most recent version. They are downloaded in JSON form and so require transformation to csv for input to phobos. The original code to do this is located in the R folder of this repository, but please be aware that this code is not up to date with MoNA's latest changes.

**Tidy and setup** 

- Make sure all inputs (experimental and theoretical) are in a standardized format.
- The MoNA data requires hydrogen manipulation. See script for details.

**Process** 

- Apply comparison function to each row (each compound @ each voltage) of the experimental data frame.
- Filter the potential matches on m/z by user-defined ppm error.
- Match polarity and column.
- Rowwise:
  MS1 (m/z) Similarity Score: exp(-0.5 * (((experimental value - theoretical value) / user-defined flexibility) ^ 2))
  MS2 cosine similarity. If both sets of MS2 values are present, calculate the similarity using the 2015 MSDial paper similarity equations.
  Total Similarity Score, according to which other similarity scores are present, also taken from 2015 MSDial paper.

Notes on CL2: 
Voltage needs to be considered since MoNA doesn't always match with the voltages we use, but the columns are still included in the download and associated data.
Filters should be adjusted for this level; too small no good matches are made, too big and the process takes a very long time (around 40 minutes).
TODO: The MoNA spreadsheets are still in the original KRH download, located in the example_data folder. The formatting code no longer works, so that part will need to be rewritten.
TODO: The total similarity score needs to be more flexible depending on what is present/absent. Messy right now.

---

### Confidence Level 3

Compounds that receive a confidence level of 3 are also considered putatively annotated, but based on fewer restrictions. phobos defines confidence level 3 as a good MS1 and z match from MoNA or KEGG.

As with MoNA, the KEGG data can found on the shared Ingalls Lab Google Drive under Collaborative_Projects --> MARS_Project --> KEGG. The public data is also available on the [KEGG](https://www.kegg.jp/kegg/) website, but requires a subscription. Additionally, the KEGG script for managing the download is out of date with KEGG's new subscription process.

Notes on Confidence Level 3:
Still adjusting the mass in MoNA to account for hydrogen.
Should probably make the cutoffs even more flexible here.
TODO: Progress bar needs to be added now that there is no MS2 similarity function.

---

### Confidence Level 4

Confidence level 4 is everything else!
